Water desalination using polyelectrolyte hydrogel: Gibbs ensemble modeling

### 2022-09-15 gibbs.tex
merged by meld the changes in intro from Lucie

\todo[inline]{
The authors can follow the following references to generally highlight the global contamination and environmental pollution problems at the beginning of the introduction part
https://doi.org/10.1016/j.jwpe.2022.102847 \cite{baaloudj2022_1}\\
https://doi.org/10.1016/j.ceramint.2022.05.151 \cite{shahzad2022}\\
https://doi.org/10.3390/catal12050500 \cite{baaloudj2022}\\
https://doi.org/10.1007/s11356-022-21160-7 \cite{guesmi2022}\\
https://doi.org/10.3390/ma15134547 \cite{kane2022}\\. 
Then, highlight the water desalination technologies and Gibbs ensemble (https://doi.org/10.3390/nano10020293 \cite{erdos2020}).
}

Two basic approaches for separating water from salt are present in modern desalination technology \cite{Miller2003,Curto2021}.

The first approach is distillation, which uses heat to cause a phase change of the water to vapor. The vapor phase is separated from the brine and condenses to liquid fresh water.
The released condensation energy is directed back to heating the feed solution.
Distillation processes were the first desalination techniques conducted on a large commercial scale and still account for a large portion of the modern world’s desalination capacity.

The second approach is to physically separate the brine components using an osmotic membrane transparent only for water molecules, which move in response to a water chemical potential difference.
In the context of our study, we will mention the reverse osmosis process (RO)~--- the major process of all the modern desalination industry, and the newly emerging membrane
technology~--- forward osmosis (FO) \cite{Akther2015}.
In RO, the difference in water chemical potential originates from a difference in pressures applied to the feed and to product solutions.
In FO, the chemical potential difference is due to an addition to solution from one side of the membrane (draw solution) the so-called draw solutes, which are lowering the water chemical potential in draw solution.

Distillation is easy and cheap technology but it is characterized by relatively high energy costs due to the dissipation of thermal energy.
On the contrary, RO uses expensive osmotic membranes, which need to be replaced regularly because of scaling and fouling.
Moreover, RO requires very high operating pressures, ranging from 20 to 200 bar, to let the water pass through the membrane.
But in terms of energy losses, it works close to the thermodynamic limit.
Thus, theoretically, per one ion pair transferred from freshwater to salty water, it consumes only the energy equal to the difference between the chemical potentials of the transfered ions.

@@

\subsection{Hydrogels for desalination}
Hydrogels are three-dimensional networks of polymer chains that are crosslinked by either physical or chemical bonds.
They are able to entrap large volumes of water attracted by the high concentration of hydrophilic groups.
When a dehydrated or deswollen hydrogel uptakes water, its polymer chains extend, creating a swelling pressure.
For example, as reported in \cite{Wack_2009} weakly crosslinked poly(acrylic acid) (PAA) copolymers with polymer volume fractions between 0.03 and 0.30 exhibit a swelling pressure ranging from 0.20--4.23 MPa.
Polyelectrolyte hydrogels, which are carrying ionic groups on the comonomer units (like PAA), are able to reject salt ions from the solution, they absorb a solution of lower salinity than that they are equilibrated with.

An important, advantageous aspect of polymer hydrogels is that they can undergo reversible volume change, \ie solution --- gel phase transitions in response to environmental stimuli.
This aspect causes hydrogels to be labeled as ‘smart’ hydrogels.
Many physical and chemical stimuli have been applied to induce various responses of such smart hydrogels, in particular, to change them from hydrophilic to hydrophobic, thereby releasing water.
The physical stimuli include: temperature, solvent composition, light, mechanical pressure, sound, electric and/or magnetic field, whilst the chemical (or biochemical) stimuli include pH, ionic strength, and specific molecular recognition \cite{Tanaka_1982,Serizawa_2001,Lietor_Santos_2009,Qiu_2001}.

The smart properties of hydrogels were utilized for desalination in a recent study by Li \etal \cite{Li2011}, where hydrogels have been considered as draw agents for FO.
In this study authors have shown, that hydrogels,
are able to absorb water across the FO membrane due to their swelling and osmotic pressure and
allow dewatering under the influence of stimuli (thermal and/or mechanical), due to their insoluble cross-linked polymer network.
For instance, Li \etal proposed the use of hydrogel made of thermoresponsive polyelectrolyte~--- copolymer of poly-N-isopropyl acrylamide (p-NIPAAm) and polyacrylic acid (p-AA).
Depending on the temperature the network of the gel appears to be either hydrophilic or hydrophobic.
In the hydrophilic state, the gel is used as a draw solute accumulating water inside, then in the hydrophobic state, the water releases out.

Moreover, from a thermodynamics perspective, the polyelectrolyte hydrogel itself is an osmotic membrane generating Donnan potential, which rejects ions between outer and inner solutions \cite{Wang_2014}.
Such a view on hydrogels was employed in a series of works by the group of prof. Wilhelm (see for example \cite{Arens_2017,Fengler_2020}).
The authors of these works proposed to get rid of the osmotic membrane and simply use only micro filtration to compress the hydrogel squeezing out the accumulated inside the gel solution.
In their method first, the deswollen hydrogel was equilibrated with saline water feed.
During equilibration, the gel was swelling absorbing water.
Then it was taken out from the feed solution and mechanically squeezed by means of a microfiltration membrane.
The squeezed out brine was obtained to have lower salinity than the feed water.

A similar approach was used by Ali \etal \cite{Ali2015}.
In their study the thermosensitive gel was used (based on copolymers p-NIPAAm and p-AA), so instead of physical compression, the dewatering was done by an external heating stimulus (sunlight).
At night the gel was equilibrating with feed water and at daytime, under sunlight, the gel was shrinking and releasing the collected inside solution.

\subsection{Physics behind the desalination process}
%The maximum entropy principle requires the density of charges to be more-less the same throughout the system medium. 
Since the polyelectrolyte gel has its own charges and its own neutralizing counterions, the density of mobile ions (which are able to freely enter and leave the gel) in gel appears to be smaller than their density outside the gel.
Therefore the internal solution of the gel has a lower density of mobile ions than that in the solution outside.
In that sense, the gel acts as an osmotic membrane separating solutions \cite{Rud2020}.
The driving force of the separation is the Donnan potential originated from the charges of the hydrogel network.
The density of mobile ions difference between ions concentrations in the internal and external solutions
is defined by Donnan law \cite{Rud2018}
\begin{equation}
\frac{\ccl\gel}{\cs}=\frac{\cs}{\cna\gel}=\sqrt{1+\left(\frac{\alpha}{2\cs\Vgel}\right)^{2}}\pm\frac{\alpha}{2\cs\Vgel}
\end{equation}
where $\ccl\gel$ and $\cna\gel$ are the internal concentration of monovalent anions and cations, $\cs$ is the salinity of external solution,  $\Vgel$ is the gel molar volume (inverse density of gel segments) and $\alpha$ is ionization degree (in the case of our study $\alpha=1$). 
The $\pm$ sign in the formula accounts for the change between the cases of gel made of polyanion or polycation.

The lower salinity solution can be extracted from the gel by means of compression and/or other stimulus providing that the charge of the gel remains constant.
In the case of gel made of weak (pH-sensitive) polyelectrolyte, the compression discharges the gel, therefore the neutralizing counterions leave out the gel diminishing the desalination effect \cite{Rud2018}. 
%Nevertheless, in\cite{Rud2018} we discussed how the release of counterions may be employed for desalination as well.

One can argue that the sole Donnan effect is insufficient to achieve a high salt rejection \cite{Cai2016} and the salinity of the water squeezed from hydrogel under a very high hydraulic pressure (up to 100 bar \cite{Fengler_2020}) turns out to be not much different from the initial.
Indeed, the use of high hydraulic pressure diminishes all the advantages of this method over RO, and the reversibility of hydrogel swelling after strong compression remains questionable. 
Nevertheless, in the current study, we model the compression of the gel limiting ourselves to low compression rates, less than 5 bar, and studied how the compression of the gel affects the surrounding salinity.
We modelled also the desalination process, as a cascade of following one by each other gel swellings and compressions driving the salinity of supernate down to drinkable water. 
%In the current study we would like to concentrate on the use of gel made of strong polyelectrolyte for water desalination. 
%Namely, to investigate how the compression of the gel affects the salinity of the surrounding the gel solution and to propose the way of water desalination employing the hydrogel swelling/deswelling set of processes.

%The target of our study is to model the thermodynamic equilibrium between the two boxes and to show how the compression of the gel affects the ion density in the outside solution.

>>>>

\todo[inline]{
The authors can follow the following references to generally highlight the global contamination and environmental pollution problems at the beginning of the introduction part
https://doi.org/10.1016/j.jwpe.2022.102847 \cite{baaloudj2022_1}\\
https://doi.org/10.1016/j.ceramint.2022.05.151 \cite{shahzad2022}\\
https://doi.org/10.3390/catal12050500 \cite{baaloudj2022}\\
https://doi.org/10.1007/s11356-022-21160-7 \cite{guesmi2022}\\
https://doi.org/10.3390/ma15134547 \cite{kane2022}\\. 
Then, highlight the water desalination technologies and Gibbs ensemble (https://doi.org/10.3390/nano10020293 \cite{erdos2020}).
}
Waste water treatment and technology is one of the greatest concerns of modern society. 
The waste water has to be disposed of both biological \cite{guesmi2022} and chemical \cite{baaloudj2022_1,shahzad2022} pollutants.
Most imporatntly, water treatment technologies are needed for the still increasing demand on production of potable water from the brine, \ie for desalination.

\subsection{Water desalination technologies}
Two basic approaches for separating water from salt are present in modern desalination technology \cite{Miller2003,Curto2021}.

The first approach is distillation, which uses heat to cause a phase change of the water to vapor. The vapor phase is separated from the brine and condenses to liquid fresh water. 
The released condensation energy is directed back to heat the feed solution.
Distillation processes were the first desalination techniques conducted on a large commercial scale and still account for a large portion of the modern world’s desalination capacity.

The second approach is to physically separate the brine components using an osmotic membrane transparent only for water molecules, which move in response to the difference in water chemical potential.
In the context of our study, we will mention the reverse osmosis process (RO) as the major process of all the modern desalination industry, and the newly emerging membrane technology~--- forward osmosis (FO) \cite{Akther2015}. 
In RO, the difference in water chemical potential originates from a difference in pressures applied to the feed and to product solutions. 
In FO, the chemical potential difference is due to an addition to solution from one side of the membrane (draw solution) the so-called draw solutes, which are lowering the water chemical potential in draw solution.

Distillation is easy and cheap technology but it is characterized by relatively high energy costs due to the dissipation of thermal energy. 
In turn, RO uses expensive osmotic membranes, which need to be replaced regularly because of scaling and fouling. 
Moreover, RO requires very high operating pressures, ranging from 20 to 200 bar, to let the water pass through the membrane. 
But, in terms of energy losses, it works close to the thermodynamic limit.
Thus, theoretically, per one ion pair transferred from freshwater to salty water, it consumes only the energy equal to the difference between the chemical potentials of the transfered ions.

@@

\subsection{Hydrogels for desalination}
Hydrogels are three-dimensional networks of polymer chains that are crosslinked by either physical or chemical bonds. 
They are able to entrap large volumes of water attracted by the high concentration of hydrophilic groups. 
When a dehydrated or deswollen hydrogel uptakes water, its polymer chains extend, creating a swelling pressure. 
For example, as reported in \cite{Wack_2009}, weakly crosslinked poly(acrylic acid) (PAA) copolymers with polymer volume fractions between 0.03 and 0.30 exhibit a swelling pressure ranging from 0.20--4.23 MPa.
Polyelectrolyte hydrogels, which are carrying ionic groups on the comonomer units (like PAA), are able to reject salt ions from the solution, they absorb a solution of lower salinity than that they are equilibrated with.

An important, advantageous aspect of polymer hydrogels is that they can undergo reversible volume change, \ie solution --- gel phase transitions in response to external stimuli. 
This aspect causes hydrogels to be labeled as ‘smart’ materials. 
Many physical and chemical stimuli have been applied to induce various responses of such smart hydrogels, in particular, to change them from hydrophilic to hydrophobic, thereby releasing water. 
The physical stimuli include: temperature, solvent composition, light, mechanical pressure, sound, electric and/or magnetic field, whilst the chemical (or biochemical) stimuli include pH, ionic strength, and specific molecular recognition \cite{Tanaka_1982,Serizawa_2001,Lietor_Santos_2009,Qiu_2001}.

Li \etal \cite{Li2011} took advantage of the smart hydrogel properties for desalination purposes. In this work, the authors used hydrogels as draw agents for FO. 
They have demonstrated that the hydrogels are able to absorb water across the FO membrane due to their swelling and osmotic pressure and
allow dewatering under the influence of stimuli (thermal and/or mechanical), due to their insoluble cross-linked polymer network.
They have proposed the use of hydrogel based on thermoresponsive polyelectrolyte~--- copolymer of poly-N-isopropyl acrylamide (p-NIPAAm) and polyacrylic acid (p-AA).
This gel network is either hydrophilic or hydrophobic, depending on the temperature.
Such gel accumulates water inside the network in its hydrophilic state, but it releases the water out in its hydrophobic state.

From a thermodynamics perspective, the polyelectrolyte hydrogel itself is an osmotic membrane generating Donnan potential, which rejects ions between outer and inner solutions \cite{Wang_2014}. 
Such a view on hydrogels was employed in a series of works by the group of prof. Wilhelm (see for example \cite{Arens_2017,Fengler_2020}). 
The authors of these works proposed to get rid of the osmotic membrane and simply use only micro filtration to compress the hydrogel squeezing out the accumulated inside the gel solution. 
In their method, the deswollen hydrogel was first equilibrated with saline water feed. 
During the equilibration, the gel was swelling, absorbing the water. 
Then it was taken out from the feed solution and mechanically squeezed by means of a microfiltration membrane. 
They have found that the brine, which was squeezed out, had lower salinity than the feed water.

Similar approach was used by Ali \etal \cite{Ali2015}. 
Here the authors used the thermosensitive gel (based on copolymers p-NIPAAm and p-AA), and instead of physical compression, the dewatering was done by an external heating stimulus (sunlight). 
The gel was equilibrated with feed water during the night, and, at daytime, under sunlight, the gel was shrinking and releasing the solution.

\subsection{Physics behind the desalination process}
Since the polyelectrolyte gel has its own charges and its own neutralizing counterions, the density of mobile ions (which are able to freely enter and leave the gel) inside the gel network appears to be smaller than their density outside the gel.
Therefore the internal solution in the gel has lower density of mobile ions than the solution outside. 
In that sense, the gel acts as an osmotic membrane separating solutions \cite{Rud2020}. 
The driving force of the separation is the Donnan potential, which originates from the charges in the hydrogel network. 
The difference between the densities (concentrations) of the mobile ions in the internal and external solutions
is defined by Donnan law \cite{Rud2018}
\begin{equation}
\frac{\ccl\gel}{\cs}=\frac{\cs}{\cna\gel}=\sqrt{1+\left(\frac{\alpha}{2\cs\Vgel}\right)^{2}}\pm\frac{\alpha}{2\cs\Vgel}
\end{equation}
where $\ccl\gel$ and $\cna\gel$ are the concentrations of monovalent anions and cations in the internal solution,
$\cs$ is the salinity of external solution,  $\Vgel$ is the gel molar volume (inverse density of gel segments)
and $\alpha$ is ionization degree (in the case of our study $\alpha=1$). 
The $\pm$ sign in the formula accounts for the sign change for the cases of polyanion or polycation gels.

As shown in \cite{Li_2011,Ali_2015,Arens_2017,Fengler_2020}, the solution of lower salinity can be extracted from the gel by means of compression and/or other stimulus, provided that the charge in the gel remains constant.
In the case of weak polyelectrolyte (pH-sensitive) gel, the compression discharges the gel and therefore the neutralizing counterions leave out the gel, diminishing the desalination effect \cite{Rud2018}.

One can argue that the sole Donnan effect is insufficient to achieve a high salt rejection \cite{Cai2016} and the salinity of the water squeezed from hydrogel under a very high hydraulic pressure (up to 100 bar \cite{Fengler_2020}) turns out to be not much different from the initial.
Indeed, the use of high hydraulic pressure diminishes all the advantages of this method over RO, and the reversibility of hydrogel swelling after strong compression remains questionable. 
Nevertheless, in this study, we model the compression of the gel limiting ourselves to low compression rates, less than 5 bar, and studied how the compression of the gel affects the surrounding salinity.
We modelled the desalination process, as a cascade of step-by-step gel swellings and compressions, driving the salinity of the supernate down to potable water.

% water desalination technologies and Gibbs ensemble (https://doi.org/10.3390/nano10020293 \cite{erdos2020}).
